A Linear Solution of the Linearity Challenge

### 2025-12-24 paper/introduction.tex
examples in subsection
removed side note

In this paper we attack the linearity challenge from a different angle:
\emph{instead of proposing new techniques that make it easy to formalize the
calculus in the challenge, we propose a (relatively) new calculus that is easy
to formalize with the existing techniques}.\Luca{dobbiamo dire perché lo spirito
della challenge è rispettato, e dunque giustificare il titolo dell'articolo.
Dire che il nostro non è un calcolo ad hoc, ma ha fondamenti logici importanti,
è completo, ed è strettamente in relazione con altri modelli analoghi (CP e
linear pi)}
%
More specifically, we describe a calculus called \LCP whose type system
coincides with the proof system of \emph{classical linear logic} and that
features \emph{linear channels} instead of sessions. Linear channels are meant
to be used for a \emph{single} communication. Structured sessions involving the
exchange of several messages can be modeled in \LCP using \emph{explicit
continuations} according to the well-known encoding of (binary) sessions into
the linear
$\pi$-calculus~\citep{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}.
%
Overall, \LCP is a linear version of \CP~\cite{Wadler14} in which sessions are
encodable instead of being featured natively.
%
These features play an important role in taming the complexity of its
formalization.
%
Using a calculus based on linear logic prevents \emph{by construction} the same
(sequential) process to own both endpoints of a session. This situation is
clearly undesirable since it does not correspond to a useful pattern of
interaction (every meaningful session requires its endpoints to be used by
parallel processes) and is a potential source of deadlocks.
%
From the standpoint of the formalization, where the representation of channel
names is a primary design choice, it spares us the need to \emph{distinguish}
the two endpoints of a session, \eg by means of
polarities~\citep{GayHole05,ZalakainDardha21} or by using different names
connected by the same binder~\citep{Vasconcelos12,GayVasconcelos25}.
%
Using a calculus with \emph{linear} channels and explicit continuations spares
us the need to \emph{update} the type of channels in typing contexts. When a
channel is used it is effectively consumed, therefore its type can be
\emph{removed} from the typing context and the type of the continuation channel
is \emph{added} back to the typing context. As is turns out, removing and adding
types is easier than updating them. Even more so considering that the type of
continuation channels can be conveniently added \emph{at the beginning} of a
typing context, since such channels are fresh by definition.

@@

The rest of the paper is organized as follows.
%
\Cref{sec:calculus} describes the syntax and the operational semantics of \LCP
and states the properties of well-typed processes that we formalize and prove,
namely typing preservation, deadlock freedom and runtime safety.
%
\Cref{sec:agda} presents a roadmap of the Agda formalization and discusses its
key aspects with particular emphasis on the representation of channel names and
of typing contexts. We assume that the reader is somewhat knowledgeable about
Agda, but we also recall a few key definitions from Agda's standard library.
%
\Cref{sec:examples} illustrates some examples of \LCP processes as they are
represented in Agda and shows how to use Agda as an interpreter for \LCP.
%
\Cref{sec:related} discusses related work more in detail by providing a
qualitative and quantitative comparison between the known formalizations of
session/linear calculi, including the one for \LCP.
%
\Cref{sec:conclusion} summarizes our contributions and discusses some ongoing
and future work.

>>>>

In this paper we attack the linearity challenge from a different angle:
\emph{instead of proposing new techniques that make it easy to formalize the
calculus in the challenge, we propose a (relatively) new calculus that is easy
to formalize with the existing techniques}.
%
More specifically, we describe a calculus called \LCP whose type system
coincides with the proof system of \emph{classical linear logic} and that
features \emph{linear channels} instead of sessions. Linear channels are meant
to be used for a \emph{single} communication. Structured sessions involving the
exchange of several messages can be modeled in \LCP using \emph{explicit
continuations} according to the well-known encoding of (binary) sessions into
the linear
$\pi$-calculus~\citep{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}.
%
Overall, \LCP is a linear version of \CP~\cite{Wadler14} in which sessions are
encodable instead of being featured natively.
%
These features play an important role in taming the complexity of its
formalization.
%
Using a calculus based on linear logic prevents \emph{by construction} the same
(sequential) process to own both endpoints of a session. This situation is
clearly undesirable since it does not correspond to a useful pattern of
interaction (every meaningful session requires its endpoints to be used by
parallel processes) and is a potential source of deadlocks.
%
From the standpoint of the formalization, where the representation of channel
names is a primary design choice, it spares us the need to \emph{distinguish}
the two endpoints of a session, \eg by means of
polarities~\citep{GayHole05,ZalakainDardha21} or by using different names
connected by the same binder~\citep{Vasconcelos12,GayVasconcelos25}.
%
Using a calculus with \emph{linear} channels and explicit continuations spares
us the need to \emph{update} the type of channels in typing contexts. When a
channel is used it is effectively consumed, therefore its type can be
\emph{removed} from the typing context and the type of the continuation channel
is \emph{added} back to the typing context. As is turns out, removing and adding
types is easier than updating them. Even more so considering that the type of
continuation channels can be conveniently added \emph{at the beginning} of a
typing context, since such channels are fresh by definition.

@@

The rest of the paper is organized as follows.
%
\Cref{sec:calculus} describes the syntax and the operational semantics of \LCP
and states the properties of well-typed processes that we formalize and prove,
namely typing preservation, deadlock freedom and runtime safety.
%
\Cref{sec:agda} presents a roadmap of the Agda formalization and discusses its
key aspects with particular emphasis on the representation of channel names and
of typing contexts. We assume that the reader is somewhat knowledgeable about
Agda, but we also recall a few key definitions from Agda's standard library.
%
% \Cref{sec:examples} illustrates some examples of \LCP processes as they are
% represented in Agda and shows how to use Agda as an interpreter for \LCP.
%
\Cref{sec:related} discusses related work more in detail by providing a
qualitative and quantitative comparison between the known formalizations of
session/linear calculi, including the one for \LCP.
%
\Cref{sec:conclusion} summarizes our contributions and discusses some ongoing
and future work.
